### 2022-10-22 dissertation/chapters/03-heuristics.tex
🚧 (chapter3): WIP - Save point

\chapter{Heuristics}
\label{chap:heuristics}

\begin{quotation}
      \textit{``It is not the strongest of the species that survives. It is also not the most intelligent that survives. It is the one that is the most adaptable to change.'' - Charles Darwin}
\end{quotation}

So far the reader has been introduced to the detailed background information on \acp{ANN}. The training process was discussed without any reference to specific optimiser algorithms to execute the training process. Specific reference is made to supervised learning in the context of \ac{FFNN}. This chapter aims to extend to this information by providing the reader with a set of optimisers that have been widely used to train \acp{ANN}. In the context of \acp{HH}, standardisation of terminology is in order and therefore these optimisers fill the role of low-level heuristics and will be referred to as heuristics from this point on.

Many different techniques have been used to train \acp{ANN} \cite{ref:kingma:2014}.  At the time of writing, the majority of work that is published around \ac{ANN} training involves the use of some gradient-based technique \cite{ref:nel:2021}. However, research has shown that these gradient-based techniques do have their disadvantages such as slow-convergence or getting trapped in local optima \cite{ref:mingguang:2009}. Other techniques have also been shown to successfully train \acp{FFNN}. These include meta-heuristics such as  \ac{PSO} \cite{ref:rakitianskaia:2012}\cite{ref:vanwyk:2014}, \ac{DE} \cite{ref:espinal:2011} and \acp{GA} \cite{ref:gupta:1999}. Various heuristics are considered for training \acp{FFNN} as the best heuristic to use is shown to be problem dependent in many cases \cite{ref:kheiri:2017}.

Although there are many different types of heuristics, the list of heuristics specified in this Chapter is by no means exhaustive. Broadly speaking this chapter focuses on two different groups of heuristics. These include classical gradient-based approaches and population-based meta-heuristics. Each technique is discussed in detail. Pseudo-code algorithms are provided and where possible and discussions follow on their advantages, disadvantages, capabilities and limitations. This subset of heuristics were selected for their characteristics and were deemed appropriate and sufficient for the problem at hand. The remainder of this chapter is structured as follows.

\begin{itemize}
      \item \textbf{Section \ref{sec:heuristics:what_is_a_heuristic}} provides a high-level definition for heuristics.

      \item \textbf{Section \ref{sec:heuristics:optimisation}} provides a brief overview of what optimisation is. It is also shown that \ac{FFNN} training is an optimisation problem.

      \item \textbf{Section \ref{sec:heuristics:gd}} presents the reader with the concept of gradient-based heuristics. The basics of \ac{BP} is provided and detailed discussions follow on seven different gradient-based low-level heuristics.

      \item \textbf{Section \ref{sec:heuristics:mh}} presents the reader with three different meta-heuristics that are show to train \acp{FFNN}.

      \item Finally, a summary of the chapter is provided in \textbf{Section \ref{sec:heuristics:summary}}.
\end{itemize}

\section{What is a heuristic?}
\label{sec:heuristics:what_is_a_heuristic}

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out, discover''. \citeauthor{ref:romanycia:1985} \cite{ref:romanycia:1985} provides a complete study on the history and origins of the term \textit{heuristic}. From their research, a proposal was made to define heuristics in the context of
\ac{AI}, as any device, be it a program, rule, piece of knowledge, which one is not entirely confident will be useful in providing a practical solution, but which one has reason to believe will be useful and which is added to a problem-solving system in expectation that on average the performance will improve.

In the context of this dissertation, the above definition holds, but can be more clearly specified such that a heuristic refers to an algorithmic search technique that serves as a guide to a search process where some good-performing candidate solution to a optimisation problem is being sought out. Different heuristics make use of different information during the search process \cite{ref:kheiri:2017}. Heuristics such as gradient-based techniques make use of the derivatives obtained by evaluating the \ac{ANN} and thus it can be said that they make use of information directly from the \textit{search space}. On the contrary, heuristics such as meta-heuristics make use of meta-information obtained as a result of evaluating the \ac{FFNN} \cite{ref:blum:2003}. This could be the ranked-performance of a population of candidate solutions, referred to as \textit{entities}. Meta-heuristics are useful when there is imperfect information about the search space \cite{ref:bianchi:2009} and are generally less problem-specific than other heuristic techniques \cite{ref:blum:2003}.

In order to understand the application of heuristics, consider first the concept of optimisation and how that relates to training \acp{FFNN}.

\section{Optimisation}
\label{sec:heuristics:optimisation}

@@

      \item A \textbf{set of unknowns or independent variables}: These affect the outcome of the objective function $f$ and is denoted as $x$. $f(x)$ is thus the quantification of the objective function over the unknowns $x$.

      \item A \textbf{set of constraints}: These restrict and limit the values that can be assigned to the unknowns $x$. Optimisation problems that contain must adhere to a set of constraints are referred to as \acp{CSP}.
\end{itemize}

The goals of heuristics is to find a set of valid values for $x$ that optimises the objective function while adhering to a set of constraints. Heuristics thus search for solutions, referred to as candidate solutions, in a set of all possible and valid solutions, referred to as the \textit{feasible search space}.

Optimisation problems come in a wide variety of forms. Mostly these functions are defined in terms of the number of variables used (uni- vs. multivariate), the number of objective functions used (single- vs. multi-objective), the degree of linearity (linear vs. quadratic/polynomial), the number of optima (uni- vs. multi-modal), the nature of the environment (static vs. dynamic), the types of variables used (separable vs. inseparable, discrete vs. continuous) and  the set of constraints that the solution has to adhere to (constrained vs unconstrained).

Optima can be defined as local or global optima. Local optima is the best optimisation of $f(x)$ in a neighbourhood of solutions, while the global optima is the best optimisation of $f(x)$ over all solutions in the solution space.

From the components of optimisation that was presented above, it can be said that the training process of an \ac{FFNN} is an optimisation problem. Consider the relationship between optimisation and supervised training of \acp{FFNN} as follows. During training, the \ac{FFNN} is evaluated by means of a loss/cost function as was presented in Chapter \ref{chap:anns}. In the context of optimisation, this loss/cost function is the object function to be minimised. For supervised learning specifically, it was mentioned that a training dataset is used to train the \ac{FFNN}. The inputs to the \ac{ANN} would be these training samples and thus serves the role of independent variables in the context of optimisation. The output of the \ac{ANN} is directly influenced by the \ac{ANN} model parameters, called the \textit{weights} of the network and is thus referred to as the dependent variables. Finally, the goal of training \acp{ANN} is to find the configuration of \textit{weights} such that the performance of the \ac{ANN} is optimal.

The following sections present the various types of heuristics that are of concern to this dissertation and the implementation of the \ac{BHH}.

\section{Gradient-Based Heuristics}
\label{sec:heuristics:gd}

Gradient-based techniques are optimisation technique that makes use of the evaluated gradients of a model in order to optimise the model. Specifically in the context of a minimisation problem, these techniques are called gradient-\textit{descent} heuristics as they minimise the loss/cost function. \Ac{GD} is generally attributed to Cauchy \cite{ref:lemarechal:2012}, who first suggested it in 1847. In 1907 Hadamard \cite{ref:hadamard:1908} independently proposed a similar method.

Although \ac{GD} techniques where not the first techniques used to train \ac{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most popular. \Ac{GD} techniques have become increasingly popular partly due to their simplicity, but mostly due to their low computational overhead compared to other techniques such as meta-heuristics and other second-order derivative methods such as Newtons' method. There are many variants of this technique, however, they all fundamentally apply the same generic \ac{GD} framework called \acl{BP}. The \ac{BP} framework is provided next and all the other gradient-based techniques that follow are built upon this idea.

\subsection{Backpropagation}
\label{sec:heuristics:gd:backpropagation}

\Ac{GD} techniques require the definition of some error function. In the context of training \acp{FFNN}, this error function is referred to as a loss/cost function. Consider the definition of the \ac{SSE} was given in Chapter\ref{algo:bhh} and is presented in Equation \ref{eq:heuristics:gd:sse} below for convenience.

\begin{equation}
      \label{eq:heuristics:gd:sse}
      \epsilon = \sum^{P_{T}}_{p=1}(t_{p} - o_{p})^{2}
\end{equation}

In Equation \ref{eq:heuristics:gd:sse}, $t_{p}$ and $o_{p}$ denote the target and ground truth output for the $p$-th training sample/pattern. $P_{t}$ is the total number of input-target vector pairs in the training set \cite{ref:engelbrecht:2007}.

The aim of \ac{GD} is then to find some candidate solution for the weights of the \ac{FFNN} that minimises the error $\epsilon$. \citeauthor{ref:engelbrecht:2007} \cite{ref:engelbrecht:2007} states that this is done by calculating the gradient of $\epsilon$ in \textit{weight-space} and to move the weight vector along the negative gradient. An illustration of this is given in Figure \ref{fig:heuristics:gd:gd_illustration}.

\begin{figure}[htbp]
      \includegraphics[width=\textwidth]{images/gradient_descent.png}
      \caption{An illustration of \ac{GD} over various timestep showing the minimisation of the error with regard to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

The update-step for \ac{GD} can then be formulated as is shown in Equations \ref{eq:heuristics:gd:update_step_part_1}-\ref{eq:heuristics:gd:update_step_part_3} below.

@@

where

@@

and

@@

In Equation \ref{eq:heuristics:gd:update_step_part_2}, $\eta$ is the learning rate. The learning rate is an important hyper-parameter in the \ac{GD} heuristic as it controls the step-size that is taken in the direction of the negative gradient at each timestep.

In the context of \ac{FFNN} training, this error signal is propagated backwards in the network. This algorithm is known as \ac{BP} and was popularised by \citeauthor{ref:werbos:1994} \cite{ref:werbos:1994}. \citeauthor{ref:nel:2021} \cite{ref:nel:2021} states that \ac{BP} provides a procedure for updating the network weights by evaluating the derivatives of the error function $\epsilon$ with respect to the weights. \Ac{BP} consists out of the two phases \cite{ref:engelbrecht:2007}. These include:

\begin{itemize}
      \item \textbf{Feedforward Pass}: During this phase the the output values if calculcated for the \ac{FFNN} for each training pattern.
      \item \textbf{Backward Propagation}: During this phase the error signal that is calculated as is shown above in Equations \ref{eq:heuristics:gd:update_step_part_1}-\ref{eq:heuristics:gd:update_step_part_3} is propagated backwards from the output layer to the input layer of the \ac{FFNN}. Weights are then adjusted as functions of the backpropagated error signal.
\end{itemize}

From the concepts mentioned above, the full \ac{BP} process for training \acp{FFNN} can be formulated. \citeauthor{ref:engelbrecht:2007} \cite{ref:engelbrecht:2007} presents such an example and is given next.

\subsubsection*{Backpropogation of Feedforward Neural Networks}

In order to present the example for \ac{BP} as applied to \acp{FFNN}, a number of assumptions are made. These include:

\begin{itemize}
      \item The \ac{SSE} is used as the objective function.
      \item The sigmoid activation function is used in both the input and output layers.
      \item Biases are included through augmented vectors as was presented in Chapter \ref{chap:anns}
      \item \Acp{SU} are used in the input and output layers.
      \item Stochastic learning is used where a small subset of the training patterns, called a \textit{mini-batch}, is presented at a time (\ac{SGD} is discussed in further detail later in Section \ref{sec:heuristics:gd:sgd}).
\end{itemize}

Then, for each training pattern $z_{p}$ the output of the objective function is given in Equation \ref{eq:heuristics:gd:bp:sse} below.

\begin{equation}
      \label{eq:heuristics:gd:bp:sse}
      \epsilon_{p} = \frac{1}{2}\left(\frac{\sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}}{K}\right)
\end{equation}

In Equation \ref{eq:heuristics:gd:bp:sse}, $K$ represents the number of output units and $t_{k,p}$ and $o_{k,p}$ are respectively the target and output values of the $k$-th output unit.

The rest of the derivations are presented only for a single pattern. The subscript $p$ is thus omitted for convenience. The output of each layer is given as

\begin{equation}
      \label{eq:heuristics:gd:bp:output_1}
      o_{k} = f_{o_{k}}(net_{o_{k}}) = \frac{1}{1 + e^{-net_{o_{k}}}}
\end{equation}

and

\begin{equation}
      \label{eq:heuristics:gd:bp:output_2}
      y_{k} = f_{y_{k}}(net_{y_{k}}) = \frac{1}{1 + e^{-net_{y_{k}}}}
\end{equation}

The weights of the \ac{FFNN} are then updated according to Equations \ref{eq:heuristics:gd:bp:output_hto} and \ref{eq:heuristics:gd:bp:output_ith} below.

\begin{equation}
      \label{eq:heuristics:gd:bp:output_hto}
      w_{kj}(t) \hspace{3mm} += \hspace{3mm} \Delta w_{kj}(t) + \alpha \Delta w_{kj}(t - 1)
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:bp:output_ith}
      v_{ji}(t) \hspace{3mm} += \hspace{3mm} \Delta v_{ji}(t) + \alpha \Delta v_{ji}(t - 1)
\end{equation}

In Equations \ref{eq:heuristics:gd:bp:output_hto} and \ref{eq:heuristics:gd:bp:output_ith} $\alpha$ represents the \textit{momentum} and is discussed in more detail later in Section \ref{sec:heuristics:gd:momentum}. The remainder of the derivations are focused on calculating $\Delta w_{kj}(t)$ and $\Delta v_{ji}(t)$. As such, the reference to time $(t)$ is omitted for convenience.

From Equation \ref{eq:heuristics:gd:bp:output_hto} it follows that

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_output}
      \frac{\partial o_{k}}{\partial net_{o_{k}}} = \frac{\partial f_{o_{k}}}{\partial net_{o_{k}}} = (1 - o_{k})o_{k} = f'_{o_{k}}
\end{equation}

and

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_hidden}
      \frac{\partial net_{o_{k}}}{\partial w_{kj}} = \frac{\partial}{\partial w_{jk}}\left( \sum^{J+1}_{j=1}w_{kj}y_{j}\right) = y_{j}
\end{equation}

In Equation \ref{eq:heuristics:gd:bp:partial_output} $f'_{o_{k}}$ represents the derivative of the corresponding activation function. From Equations \ref{eq:heuristics:gd:bp:partial_output} and \ref{eq:heuristics:gd:bp:partial_hidden}, using the chain-rule, it follows that

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_output_hidden}
      \begin{split}
            \frac{\partial o_{k}}{\partial w_{kj}}
            &= \frac{\partial o_{k}}{\partial net_{o_{k}}}\frac{\partial net_{o_{k}}}{\partial w_{kj}}\\
            &= (1 - o_{k})o_{k}y_{j}\\
            &= f'_{o_{k}}y_{j}
      \end{split}
\end{equation}

From Equation \ref{eq:heuristics:gd:bp:sse} it follows that

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_error}
      \frac{\partial \epsilon}{\partial o_{k}} = \frac{\partial}{\partial o_{k}}\left( \frac{1}{2} \sum^{K}_{k=1}(t_{k} - o_{k})^{2} \right) = -(t_{k} - o_{k})
\end{equation}

Define the error signal that needs to be back-propagated as $\delta_{o_{k}} = \frac{\partial \epsilon}{\partial net_{o_{k}}}$. Then, from Equations \ref{eq:heuristics:gd:bp:partial_output} and \ref{eq:heuristics:gd:bp:partial_error} it follows that

\begin{equation}
      \label{eq:heuristics:gd:bp:delta_output}
      \begin{split}
            \delta_{o_{k}}
            &= \frac{\partial \epsilon}{\partial net_{o_{k}}}\\
            &= \frac{\partial \epsilon}{\partial o_{k}}\frac{\partial o_{k}}{\partial net_{o_{k}}}\\
            &= -(t_{k} - o_{k})(1 - o_{k})o_{k}\\
            &= -(t_{k} - o_{k})f'_{o_{k}}
      \end{split}
\end{equation}

The changes in the hidden-to-output weights are then computed from Equations \ref{eq:heuristics:gd:bp:partial_output_hidden}, \ref{eq:heuristics:gd:bp:partial_error} and \ref{eq:heuristics:gd:bp:delta_output} as follows.

\begin{equation}
      \label{eq:heuristics:gd:bp:delta_hidden_to_output}
      \begin{split}
            \Delta w_{kj}
            &= \eta \left( - \frac{\partial \epsilon}{\partial w_{kj}} \right)\\
            &= -\eta \frac{\partial \epsilon}{\partial o_{k}}\frac{\partial o_{k}}{\partial w_{kj}}\\
            &= -\eta \delta_{o_{k}}y_{j}
      \end{split}
\end{equation}

Continuing on with the input-to-hidden weights as follows

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_input_to_hidden_part_1}
      \begin{split}
            \frac{\partial y_{j}}{\partial net_{y_{j}}}
            &= \frac{\partial f_{y_{j}}}{\partial net_{y_{j}}}\\
            &= (1 - y_{j})y_{j}\\
            &= f'_{y_{j}}
      \end{split}
\end{equation}

and

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_input_to_hidden_part_2}
      \begin{split}
            \frac{\partial net_{y_{j}}}{\partial v_{ji}}
            &= \frac{\partial}{\partial v_{ji}}\left( \sum^{I+1}_{i=1}v_{ji}z_{i}\right)\\
            &= z_{i}
      \end{split}
\end{equation}

From Equations \ref{eq:heuristics:gd:bp:partial_input_to_hidden_part_1} and \ref{eq:heuristics:gd:bp:partial_input_to_hidden_part_2} it follows that

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_input_to_hidden_part_3}
      \begin{split}
            \frac{\partial y_{j}}{\partial v_{ji}}
            &= \frac{\partial y_{j}}{\partial net_{y_{j}}}\frac{\partial net_{y_{j}}}{\partial v_{ji}}\\
            &= (1 - y_{j})y_{j}z_{i}\\
            &= f'_{y_{j}}z_{i}
      \end{split}
\end{equation}

and

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_input_to_hidden_part_4}
      \begin{split}
            \frac{\partial net_{o_{k}}}{\partial y_{j}}
            &= \frac{\partial}{\partial y_{j}}\left( \sum^{J+1}_{j=1}w_{kj}y_{j}\right)\\
            &= w_{kj}
      \end{split}
\end{equation}

From Equations \ref{eq:heuristics:gd:bp:delta_hidden_to_output} and \ref{eq:heuristics:gd:bp:partial_input_to_hidden_part_4} it follows that

\begin{equation}
      \label{eq:heuristics:gd:bp:partial_input_to_hidden_part_5}
      \begin{split}
            \frac{\partial \epsilon}{\partial y_{j}}
            &= \frac{\partial}{\partial y_{j}}\left( \frac{1}{2} \sum^{K}_{k=1} (t_{k} - o_{k})^{2} \right)\\
            &= \sum^{K}_{k=1}\frac{}{} \frac{\partial \epsilon}{\partial o_{k}} \frac{\partial o_{k}}{\partial net_{o_{k}}} \frac{\partial net_{o_{k}}}{\partial y_{j}}\\
            &= \sum^{K}_{k=1} \frac{\partial \epsilon}{\partial net_{o_{k}}} \frac{\partial net_{o_{k}}}{\partial y_{j}}\\
            &= \sum^{K}_{k=1} \delta_{o_{k}}w_{kj}
      \end{split}
\end{equation}

Define the hidden layer error which needs to be back-propagated from Equations \ref{eq:heuristics:gd:bp:partial_input_to_hidden_part_1} and \ref{eq:heuristics:gd:bp:partial_input_to_hidden_part_5} as

\begin{equation}
      \label{eq:heuristics:gd:bp:delta_hidden}
      \begin{split}
            \delta_{y_{j}}
            &= \frac{\partial \epsilon}{\partial net_{y_{j}}}\\
            &= \frac{\partial \epsilon}{\partial y_{j}} \frac{\partial y_{j}}{\partial net_{y_{j}}}\\
            &= \sum^{K}_{k=1}\delta_{o_{k}}w_{kj}f'_{y_{j}}
      \end{split}
\end{equation}

Finally, the changes to the input-to-hidden weights can then be calculated from Equations \ref{eq:heuristics:gd:bp:partial_input_to_hidden_part_3}, \ref{eq:heuristics:gd:bp:partial_input_to_hidden_part_5} and \ref{eq:heuristics:gd:bp:delta_hidden} as follows

\begin{equation}
      \label{eq:heuristics:gd:bp:delta_input_to_hidden}
      \begin{split}
            \Delta v_{ji}
            &= \eta \left( - \frac{\partial \epsilon}{\partial v_{ji}} \right)\\
            &= -\eta \frac{\partial \epsilon}{\partial y_{j}} \frac{\partial y_{j}}{\partial v_{ji}}\\
            &= -\eta \delta_{y_{j}}z_{i}
      \end{split}
\end{equation}

If the \ac{FFNN} include direct weights from the input to output layer, the following additional weight updates are required.

\begin{equation}
      \label{eq:heuristics:gd:bp:delta_input_to_output}
      \begin{split}
            \Delta u_{ki}
            &= \eta \left( - \frac{\partial \epsilon}{\partial u_{ki}} \right)\\
            &= -\eta \frac{\partial \epsilon}{\partial o_{k}} \frac{\partial o_{k}}{\partial u_{ki}}\\
            &= -\eta \delta_{o_{k}}z_{i}
      \end{split}
\end{equation}

In Equation \ref{eq:heuristics:gd:bp:delta_input_to_output} above, $u_{ki}$ is a weight from the $i$-th input unit to the $k$-th output unit.

The derivations above assumed stochastic learning, however, if batch learning is applied, weight updates are given in Equations \ref{eq:heuristics:gd:bp:delta_hidden_to_output_batch_learning} and \ref{eq:heuristics:gd:bp:delta_input_to_hidden_batch_learning} below

\begin{equation}
      \label{eq:heuristics:gd:bp:delta_hidden_to_output_batch_learning}
      \begin{split}
            \Delta w_{kj}(t) = \sum^{P_{T}}_{p=1} \Delta w_{kj,p}(t)
      \end{split}
\end{equation}

\begin{equation}
      \label{eq:heuristics:gd:bp:delta_input_to_hidden_batch_learning}
      \begin{split}
            \Delta v_{kj}(t) = \sum^{P_{T}}_{p=1} \Delta v_{kj,p}(t)
      \end{split}
\end{equation}

where $\Delta w_{kj}(t)$ and $\Delta v_{kj}(t)$ are weight updates for individual patterns $p$ and $P_{T}$ is the total number of patterns in the training set.

Equations \ref{eq:heuristics:gd:bp:sse} through \ref{eq:heuristics:gd:bp:delta_input_to_hidden} provided the reader with the detailed derivations of the \ac{BP} weight-update components as it resolves from the error signal. The following sections provide the reader with variants of the \ac{BP} algorithm as presented in this section.

\subsection{Stochastic Gradient Descent}
\label{sec:heuristics:gd:sgd}

Section \ref{sec:heuristics:gd:backpropagation} already assumed the use of stochastic learning and thus provided the detailed derivation of a specific variant of \ac{BP} called \acl{SGD}. Separation of theory and implementation is done such that the detail of the theory need not be repeated for each gradient-based variant. This section shines light on the algorithmic implementation of \ac{BP} with specific context to \ac{SGD}. The algorithm is provided followed by a simplification of the update-step that provides the basis for the next gradient-based heuristics that follow.

The pseudo-code implementation for the \ac{SGD} algorithm is taken from \cite{ref:engelbrecht:2007} and is presented in Algorithm \ref{algo:heuristics:gd:sgd} below.

@@

In Equation \ref{eq:heuristics:gd:sgd} $w$ refers to the weights of the \ac{FFNN}, $\eta$ refers to the learning rate as before and $\nabla_{w}\mathcal{E}(w)$ refers to the gradient of the loss/cost function w.r.t the output of the \ac{FFNN} as a result of its weights. Note that all subscripts w.r.t. layers and training patterns are omitted for convenience.

>>>>

\chapter{Heuristics}\label{chap:heuristics}

\begin{quotation}
      \noindent ``It is not the strongest of the species that survives. It is also not the most intelligent that survives. It is the one that is the most adaptable to change.''
\end{quotation}
\begin{flushright}
      Charles Darwin
\end{flushright}

\noindent
Many different techniques have been used to train \acp{FFNN} \cite{ref:kingma:2014}. Finding the best technique to use to train an \acs{FFNN} has been shown to be problem dependent in many cases \cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques \cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima \cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including meta-heuristics such as \acf{PSO} \cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE} \cite{ref:espinal:2011} and \acfp{GA} \cite{ref:gupta:1999}.

Chapter~\ref{chap:anns} briefly introduced the reader to the concept of \index{heuristic}\index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. This chapter presents more detailed background information on various different \index{heuristic}heuristics that have been used to train \acp{FFNN}. Broadly speaking, this chapter focuses on two different groups of \index{heuristic}heuristics, including classical gradient-based approaches and population-based \index{meta-heuristic}meta-heuristics. Each technique is presented and discussed in detail. Pseudo-code algorithms are provided for each technique and discussions follow on advantages, disadvantages, capabilities and limitations. The remainder of this chapter is structured as follows.

\begin{itemize}
      \item \textbf{Section~\ref{sec:heuristics:optimisation}} provides a brief review of optimisation. It is shown that training of \acp{FFNN} is an optimisation problem.

      \item \textbf{Section~\ref{sec:heuristics:what_is_a_heuristic}} provides background information on the origins and definition the term \index{heuristic}\textit{heuristic}. It is shown that \index{heuristic}heuristics are a class of algorithms that are used to solve optimisation problems.

      \item \textbf{Section~\ref{sec:heuristics:gd}} presents the reader with seven low-level, gradient-based \index{heuristic}heuristics, including \acf{SGD}, \acf{Momentum}, \acf{NAG}, \acf{Adagrad}, \acf{RMSProp}, \acf{Adadelta} and \acf{Adam}.

      \item \textbf{Section~\ref{sec:heuristics:mh}} presents the reader with three different population-baed \index{meta-heuristic}meta-heuristics, including \acf{PSO}, \acf{DE} and \acfp{GA}.

      \item \textbf{Section~\ref{sec:heuristics:summary}} provides the reader with a brief summary of the chapter.
\end{itemize}

\section{Optimisation}\label{sec:heuristics:optimisation}

@@

      \item A \textbf{set of unknowns or independent variables}: Affects the outcome of the objective function $f$ and is denoted as $x$. $f(x)$ is thus the quantification of the objective function over the unknowns, represented by $x$. Note that $x$ could be a scalar value, a vector or a matrix and notation is left out for simplicity.

      \item A \textbf{set of constraints}: Restrict and limit the values that can be assigned to the unknowns, represented by $x$. Optimisation problems that must adhere to a set of constraints are referred to as \acfp{CSP}.
\end{itemize}

\noindent
Optimisation problems come in a wide variety, and can be defined in terms of the number of variables used (uni- vs. multivariate), the number of objective functions used (single- vs. multi-objective), the degree of linearity (linear vs. quadratic/polynomial), the number of optima (uni- vs. multi-modal), the nature of the environment (static vs. dynamic), the types of variables used (separable vs. inseparable, discrete vs. continuous) and the set of constraints that the solution must adhere to (constrained vs unconstrained).

Optima can be defined as \textit{local} or \textit{global} optima. Local optima is the best optimisation of $f(x)$ in a neighbourhood of solutions, while the global optima is the best optimisation of $f(x)$ over all solutions in the solution space.

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \ac{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sough out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic. Discussions on the details of heuristics are presented in the following sections.

\section{What is a heuristic?}\label{sec:heuristics:what_is_a_heuristic}

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''. \citeauthor{ref:romanycia:1985} \cite{ref:romanycia:1985} provides a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \ac{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process \cite{ref:kheiri:2017}. During the training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \ac{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \index{hmeta-euristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \ac{FFNN} \cite{ref:blum:2003}. The meta-information that \index{meta-heuristic}meta-heuristics make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space \cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics \cite{ref:blum:2003}. This disseration takes a particular interest in gradient-based \index{heuristic}heuristics and \index{meta-heuristic}meta-heuristics. Each of these are presented in detail in the following sections.

\section{Gradient-Based Heuristics}\label{sec:heuristics:gd}

Gradient-based \index{heuristic}heuristics are optimisation techniques that make use of derivates obtained from evaluating the model being optimised. Specifically, in the context of a minimisation problem, these techniques are called \acf{GD} heuristics as they \textit{minimise} some loss function. \Acs{GD} is generally attributed to Cauchy \cite{ref:lemarechal:2012}, who first suggested it in 1847. In 1907 Hadamard \cite{ref:hadamard:1908} independently proposed a similar method.

Although \acs{GD} \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \ac{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most whidely used. \Acs{GD} heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \index{meta-heuristic}meta-heuristics and other second-order derivative methods such as Newtons' method. There are many variants of \acs{GD} \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acs{GD} framework called \acl{BP}.

\subsection{Backpropagation}\label{sec:heuristics:gd:backpropagation}

Chapter~\ref{chap:anns} introduced the reader to \index {supervised learning}supervised learning and presented a number of loss functions. In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \ac{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \ac{FFNN} in order to minimise the error \cite{ref:engelbrecht:2007}. \citeauthor{ref:engelbrecht:2007} \cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient. An illustration of \acs{GD} is given in Figure~\ref{fig:heuristics:gd:gd_illustration}.

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.85\textwidth]{images/gradient_descent.pdf}
      \caption{An illustration of \ac{GD} over various timesteps showing the minimisation of the error with regards to weight value.}
      \label{fig:heuristics:gd:gd_illustration}
\end{figure}

\noindent
The update step for \ac{GD} can be formulated as is shown in Equation~\eqref{eq:heuristics:gd:update_step_part_1} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_1} above, the delta weight $\Delta w$, at index $i$ and timestep $t$ is given by Equation~\eqref{eq:heuristics:gd:update_step_part_2} below.

@@

\noindent
In Equation~\eqref{eq:heuristics:gd:update_step_part_2} above, $\eta$ is the learning rate. The learning rate is an important hyper-parameter in gradient-based \index{heuristic}heuristic as it controls the step-size that is taken in the direction of the negative gradient at each timestep. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight $w$, at index $i$, is given by Equation~\eqref{eq:heuristics:gd:update_step_part_3} below.

@@

\noindent
In the context of training \textit{shallow} \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer. The algorithm that propagates the error signal backwardds is known as \acf{BP}. \Acs{BP} was popularised by \citeauthor{ref:werbos:1994} \cite{ref:werbos:1994}. \citeauthor{ref:nel:2021} \cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\epsilon$ with respect to the weights at each layer. \citeauthor{ref:engelbrecht:2007} \cite{ref:engelbrecht:2007}describes the \acs{BP} process in two steps:

\begin{itemize}
      \item \textbf{Feedforward Pass}: During this phase the output values of the \acs{FFNN} is calculcated for each training pattern.
      \item \textbf{Backward Propagation}: During this phase the error signal that is calculated as is shown above in Equations~\eqref{eq:heuristics:gd:update_step_part_1}-~\eqref{eq:heuristics:gd:update_step_part_3} is propagated backwards from the output layer, through the hidden layer, to the input layer of the \ac{FFNN}. Weights are then adjusted as functions of the backpropagated error signal.
\end{itemize}

\subsection{Stochastic Gradient Descent}\label{sec:heuristics:gd:sgd}

This section shines light on the algorithmic implementation of \ac{BP} with specific context to \ac{SGD}. The algorithm is provided followed by a simplification of the update-step that provides the basis for the next gradient-based heuristics that follow.

The pseudo-code implementation for the \ac{SGD} algorithm is taken from \cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:sgd} below.

@@

In Equation~\ref{eq:heuristics:gd:sgd} $w$ refers to the weights of the \ac{FFNN}, $\eta$ refers to the learning rate as before and $\nabla_{w}\mathcal{E}(w)$ refers to the gradient of the loss/cost function w.r.t the output of the \ac{FFNN} as a result of its weights. Note that all subscripts w.r.t. layers and training patterns are omitted for convenience.
